Mammalian protein-coding genes exhibit widespread beneficial mutations that are not adaptive

### 2023-05-02 manuscript/scifile.tex
Prepare submission tex files. Updated figures.

%! BibTeX Compiler = biber
%TC:ignore
\documentclass{article}

\usepackage{xcolor, colortbl}
\definecolor{BLUELINK}{HTML}{0645AD}
\definecolor{DARKBLUELINK}{HTML}{0B0080}
\definecolor{LIGHTGREY}{gray}{0.9}
\PassOptionsToPackage{hyphens}{url}
\usepackage[colorlinks=false]{hyperref}
% for linking between references, figures, TOC, etc in the pdf document
\hypersetup{colorlinks,
    linkcolor=DARKBLUELINK,
    anchorcolor=DARKBLUELINK,
    citecolor=DARKBLUELINK,
    filecolor=DARKBLUELINK,
    menucolor=DARKBLUELINK,
    urlcolor=BLUELINK
} % Color citation links in purple
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{naturalnames}{hyperref}

\usepackage{biorxiv}
\usepackage[backend=biber,eprint=false,isbn=false,url=false,intitle=true,style=nature,date=year]{biblatex}
\addbibresource{codon_models.bib}

\usepackage{url}
\usepackage{amssymb,amsfonts,amsmath,amsthm,mathtools}
\usepackage{lmodern}
\usepackage{xfrac, nicefrac}
\usepackage{bm}
\usepackage{listings, enumerate, enumitem}
\usepackage[export]{adjustbox}
\usepackage{graphicx}
\usepackage{bbold}
\usepackage{pdfpages}
\pdfinclusioncopyfonts=1
\usepackage{lineno}
\usepackage{tabu}
\usepackage{hhline}
\usepackage{multicol,multirow,array}
\usepackage{etoolbox}
\usepackage{booktabs}
\usepackage{makecell}
\usepackage{orcidlink}

\newcommand{\NS}[1]{\textcolor{red}{\textbf{\emph{[NS: #1]}}}}

\newcommand{\UniDimArray}[1]{\bm{#1}}
\newcommand{\der}{\text{d}}
\newcommand{\e}{\text{e}}
\newcommand{\Ne}{N_{\text{e}}}
\newcommand{\proba}{\mathbb{P}}
\newcommand{\pfix}{\proba_{\text{fix}}}
\newcommand{\dn}{d_N}
\newcommand{\ds}{d_S}
\newcommand{\dnds}{\dn / \ds}
\newcommand{\Sphy}{S_{0}}
\newcommand{\SphyDel}{\mathcal{D}_0}
\newcommand{\SphyNeu}{\mathcal{N}_0}
\newcommand{\SphyBen}{\mathcal{B}_0}
\newcommand{\Sphyclass}{x}
\newcommand{\SphyclassAlt}{y}
\newcommand{\given}{\mid}
\newcommand{\Spop}{S}
\newcommand{\SpopDel}{\mathcal{D}}
\newcommand{\SpopNeu}{\mathcal{N}}
\newcommand{\SpopBen}{\mathcal{B}}
\newcommand{\ProbaPopDel}{\proba [ \SpopDel]}
\newcommand{\ProbaPopNeu}{\proba [ \SpopNeu ]}
\newcommand{\ProbaPopBen}{\proba [ \SpopBen ]}
\newcommand{\AdvMean}{\beta_b}
\newcommand{\DelMean}{\beta_d}
\newcommand{\thetaSyn}{\theta_{\text{S}}}
\renewcommand{\baselinestretch}{1.5}
\renewcommand{\arraystretch}{0.6}
\linenumbers
\frenchspacing

\title{Mammalian protein-coding genes exhibit widespread beneficial mutations that are not adaptive}
\rhead{\scshape Beneficial yet non-adaptive mutations}

\author{
    \large
    \textbf{T. {Latrille}$^{1\dag}$\orcidlink{0000-0002-9643-4668}, J. {Joseph}$^{2\dag}$, D.~A. {Hartasánchez}$^{1}$\orcidlink{0000-0003-2596-6883}, N. {Salamin}$^{1}$\orcidlink{0000-0002-3963-4954}}\\
    \normalsize
    $^{1}$Department of Computational Biology, Université de Lausanne, Lausanne, Switzerland\\
    $^{2}$Laboratoire de Biométrie et Biologie Evolutive, UMR5558, Université Lyon 1, Villeurbanne, France \\
    $^{\dag}$These authors contributed equally to this work\\
    \texttt{\href{mailto:thibault.latrille@ens-lyon.org}{thibault.latrille@ens-lyon.org}} \\
}

% The submission checklist is available at:
% https://docs.google.com/document/d/1aHCF1on0mHTK2DoroPCLP_xto4xwHnGJCGwaL1Vm2UE/edit?usp=sharing

\begin{document}
    \maketitle

%TC:endignore
    \begin{abstract}
        Mutations can be beneficial by bringing innovation to their bearer, allowing them to adapt to environmental change.
        However, mutations can also be beneficial because they are reverting previous deleterious changes, simply restoring pre-existing functions.
        By integrating phylogenomic and population data in mammals, we estimated the contribution of beneficial back-mutations to molecular evolution, which had remained widely overlooked so far.
        Applying a mutation-selection model, we estimated amino acid fitness landscapes for mammalian protein-coding genes.
        Assuming these fitness landscapes do not change, any beneficial mutation is necessarily a back-mutation.
        We confirmed that these beneficial back-mutations are positively selected in extant populations and demonstrated that a substantial part of ongoing positive selection is not driven by adaptation to environmental change.
    \end{abstract}

    \keywords{Adaptation \and Back-mutations \and Phylogenetics \and Population-genetics \and Codon models }

    Adaptation is one of the main processes shaping the diversity of forms and functions across the tree of life~\cite{darwin_origin_1859}.
    Evolutionary adaptation is tightly linked to environmental change and species responding to this change~\cite{merrell_adaptive_1994, gavrilets_adaptive_2009}.
    For adaptation to occur, there must be variation within populations, which mostly appears via mutations in the DNA sequence.
    While neutral mutations will not impact an individual's fitness, deleterious mutations have a negative effect, and beneficial mutations improve their bearer's fitness.
    A beneficial mutation is thus more likely than a neutral mutation to invade the population and reach fixation, resulting in a substitution at the species level.
    Upon environmental change, because adaptive beneficial mutations toward new fitness optima are more likely, the number of substitutions also increases (fig.~\ref{fig:fitness-landscape}A).
    An increased substitution rate is thus commonly interpreted as a sign of adaptation~\cite{mcdonald_adaptative_1991, smith_adaptive_2002, welch_estimating_2006}.
    The availability of large-scale genomic data and the development of theoretical models have enabled the detection and quantification of substitution rate changes across genes and lineages~\cite{yang_statistical_2000, eyre-walker_genomic_2006, moutinho_variation_2019}.
    These approaches, now common practice in evolutionary biology, helped better understand the processes underpinning the rates of molecular evolution.
    However, a collateral effect has been conflating beneficial mutations with adaptive evolution when adaptive evolution is not the only process that can lead to beneficial mutations~\cite{charlesworth_other_2007, mustonen_fitness_2009}.

@@

    %In this study, we seek empirical evidence for beneficial back-mutations and test long-standing related questions, such as how common they are and what the probability is for any beneficial mutation observed in current populations to be a back-mutation.
    %By addressing these questions, we provide novel insight into the process of adaptation and, more broadly, into how natural selection has affected protein-coding DNA sequences throughout evolution.
    \subsection*{Fitness landscape reconstruction}
    The mutation-selection framework permits to link the patterns of substitution along a phylogeny with the underlying fitness landscape~\cite{halpern_evolutionary_1998, mccandlish_modeling_2014}.
    Such mutation-selection models applied to protein-coding DNA sequences allow us to estimate relative fitnesses for all amino acids for each site of the sequence, explicitly assuming that the underlying fitness landscape is stable along the phylogenetic tree~\cite{rodrigue_mechanistic_2010, tamuri_estimating_2012, rodrigue_detecting_2017}.
    Importantly, because mutation-selection models at the phylogenetic scale are based on population-genetics equations, their estimates of selection coefficients are directly interpretable as fitness effects for an individual within a population.
    %These methods account for phylogenetic relationships, the structure of the genetic code and accommodate for underlying mutational biases~\cite{rodrigue_mechanistic_2010, tamuri_estimating_2012, latrille_improved_2022}.

@@

    This result further indicated that using $\dnds$ as an estimate of adaptation is biased due to the presence of beneficial back-mutations among the non-synonymous substitutions.
    By discarding all beneficial back-mutations we can obtain an estimate of $\dnds$ which is not inflated.
    By comparing these two ways of calculating $\dnds$ (see section~\ref{subsec:substitution-mapping-in-the-terminal-branch} in Materials \& Methods), we calculated that beneficial back-mutations inflate $\dnds$ values by between 9 and 11\% across genera while only representing between 1.2 and 1.4\% of non-synonymous mutations (table~S2-3 and fig.~\ref{fig:homo-afr-results}A for humans).

@@

    In addition to clinical prognosis, frequencies at which SNPs are segregating within populations provide information on their selective effects.
    For instance, deleterious SNPs usually segregate at lower frequencies because of purifying selection, which tends to remove them from the population (fig.~\ref{fig:homo-afr-results}C for humans).
    By gathering information across many SNPs, it is possible to estimate the distribution of fitness effects (DFE) at the population scale, taking synonymous SNPs as a neutral expectation~\cite{eyre-walker_distribution_2006, eyre-walker_estimating_2009, galtier_adaptive_2016, tataru_inference_2017}.
    From the estimated DFE, we can derive the proportion of beneficial mutations ($\ProbaPopBen$), nearly-neutral mutations ($\ProbaPopNeu$) and deleterious mutations ($\ProbaPopDel$) at the population scale (see section~\ref{subsec:s-polymorphism-method} in Materials \& Methods).
    These approaches offer a unique opportunity to contrast selection coefficients estimated at the population scale ($\Spop$) and at the phylogenetic scale ($\Sphy$).

    Across our selection classes ($\SphyDel$, $\SphyNeu$ and $\SphyBen$), one can ultimately estimate the proportion of correct and incorrect predictions, leading to an estimation of \textit{precision} and \textit{recall} (fig.~\ref{fig:method}F-G and section~\ref{subsec:precisison_recall} in Materials \& Methods).
    Across 28 populations of different mammal species, mutations predicted to be deleterious at the phylogenetic scale ($\ProbaPopDel$) were indeed purged at the population scale, with a \textit{precision} in the range of 89--96\% (table~\ref{table:proba} and fig.~\ref{fig:homo-afr-results}D for humans).
    Conversely, a \textit{recall} in the range of 95--100\% implies that mutations found to be deleterious at the population scale were most likely also predicted to be deleterious at the phylogenetic scale (table~\ref{table:proba}).
    Altogether, purifying selection is largely predictable and amino acids with negative fitness across mammals have been effectively purged away in each population.

@@

    %We have to acknowledge that the mammalian case study might also not be representative of other clades.
    %Indeed, long-term population sizes are relatively small in mammals, which allow large amounts of deleterious mutations to be fixed, and thus, a lot of opportunities for beneficial back-mutations to arise.
    %It would thus be of great interest to reproduce our analysis with other clades, such as \textit{Drosophila} or \textit{Saccharomyces} which have higher effective population sizes.
    %Indeed when the effective population size is higher, the number of effectively neutral mutational opportunities should be reduced.
    %Allowing for changes in effective population sizes along the phylogeny have been developed~\cite{latrille_inferring_2021}, but are, for the moment, too computationally intensive to be performed genome-wide.
    %This reduction should mechanistically increase the proportion of adaptive substitutions.
    %However, this does not imply that the rate of adaptation (i.e the number of changes in the fitness landscape per site per generation) is higher~\cite{galtier_adaptive_2016}.

@@

    %Regarding the distribution of fitness effects (DFE) estimated at the population scale, there is no consensus on its expected shape~\cite{welch_divergence_2008, bataillon_effects_2014}, although it has been found to be reliably constant between closely related species~\cite{castellano_comparison_2019}.
    %We modelled the DFE as a continuous distribution, but to check for robustness in our results despite the uncertainty of the underlying DFE, we also performed our analysis assuming a discretized DFE.
    %Our results are consistent, with the proportion of back-mutations among all beneficial mutations ranging from XX to XX\% across populations when modelling a DFE as a discrete distributio.
    %Consequently, even though our general framework contrasting the phylogenetic and population scales is theoretically consistent, we acknowledge that estimating the proportion of back-mutations among all beneficial ones is dependent on the the underlying DFE at the population scale and the threshold used to define a beneficial mutation.
    %As a result, further work is required to investigate the shape of the DFE at the population scale.
    %Notably, the reconstructed distribution of $\Sphy$ values obtained in our study has a negative mode (around $\Sphy=-5$), consistent with experimental DFEs obtained by mutation accumulation.
    %Contrarily, continuous DFEs estimated at the population scale have by definition a mode at 0.
    %We argue that methods estimating DFE at the population scale should assume a mode shifted toward negative values of $\Spop$.

@@

    %\subsection*{The role of epistasis in mammalian protein-coding genes}
    %
    %In our study, we focused on mutations toward a more optimal amino-acid at a given site, assuming that amino-acid fitness landscapes are site-specific and also independent of one another.
    %Contrarily, under epistasis, the fitness effect of a mutation at a particular site depends on the amino-acids present at other sites.
    %Epistasis therefore allows for compensatory mutations, which restore fitness through mutations at loci different from where deleterious mutations take place, and this process also leaves signals interpretable as non-adaptive beneficial mutations~\cite{hartl_compensatory_1996, pollock_strong_2014, starr_epistasis_2016}.
    %Also, epistasis has been shown to play a role in the evolution of protein-coding genes, with amino-acid residues in contact within a protein or between proteins tending to co-evolve~\cite{breen_epistasis_2012, starr_epistasis_2016}.
    %Particularly, the residues in contact co-evolve to become more compatible with each other generating an entrenchment, making empirical evidence for beneficial back-mutations harder to find since restoring mutations becomes less and less likely to occur over time~\cite{goldstein_nonadaptive_2015, goldstein_sequence_2017, park_epistatic_2022}.
    %Nevertheless, the fact that we observe such a high proportion of beneficial back-mutations suggests that the underlying assumptions of our model, namely site-independence, implying no epistasis, and a static fitness landscape, are a reasonable approximation for the underlying fitness landscape of proteins.
    %Our results imply that the fitness effects of new mutations are mostly conserved across mammalian orthologs, in agreement with other studies showing that for conserved orthologs with similar structures and functions, models without epistasis provide a reasonable estimate of fitness effects in protein coding genes~\cite{ashenberg_mutational_2013, youssef_consequences_2020}.

@@

%TC:ignore
    \section*{Acknowledgements}
    \label{sec:acknowledgment}
    We gratefully acknowledge the help of Mélodie Bastian, Nicolas Lartillot, Carina Farah Mugal, Laurent Duret and Nicolas Gambardella for their advice and reviews concerning this manuscript.
    This work was performed using the computing facilities of the CC LBBE/PRABI\@.
    This study makes use of data generated by the NextGen Consortium.
    \textbf{Funding:}
    Université de Lausanne; Agence Nationale de la Recherche, Grant ANR-19-CE12-0019 / HotRec.
    The European Union’s Seventh Framework Programme (FP7/2010-2014) provided funding for the project under grant agreement no 244356 - “NextGen”.
    \textbf{Author contributions:}
    Original idea: T.L.\ and J.J.;
    Model conception: T.L., J.J.\ and N.S.;
    Code: T.L.;
    Data analyses: T.L.\ and J.J.;
    Interpretation: T.L., J.J., D.A.H.\ and N.S.;
    First draft: T.L.\ and J.J.;
    Editing and revisions: T.L., J.J., D.A.H.\ and N.S.
    Project management and funding: N.S\@.
    \textbf{Competing interests:}
    The authors declare no conflicts of interest.
    \textbf{Data and materials availability:}
    The data underlying this article are available at \url{https://doi.org/10.5281/zenodo.7878954}.
    Snakemake pipeline, analysis scripts and documentation are available at \href{https://github.com/ThibaultLatrille/SelCoeff}{github.com/ThibaultLatrille/SelCoeff}.

    \section{Materials \& Methods}
    \label{sec:methods}

    \subsection{Phylogenetic dataset}\label{subsec:phylo-dataset}

    Protein-coding DNA sequence alignments in placental mammals and their corresponding gene trees come from the \href{https://www.orthomam.univ-montp2.fr}{OrthoMaM} database and were processed as in \textcite{latrille_genes_2023}.
    OrthoMaM contains a total of $116$ mammalian reference sequences in v10c~\cite{ranwez_orthomam_2007, douzery_orthomam_2014, scornavacca_orthomam_2019}.
    Genes located on the X and Y chromosomes and on the mitochondrial genome were discarded from the analysis because the level of polymorphism –~which is necessary for population-based analyses~– is expected to be different in these three regions compared to the autosomal genome.
    Sequences of species for which we used population-level polymorphism (see section \ref{subsec:polymorphism-dataset}) and their sister species, were removed from the analysis to ensure independence between the data used in the phylogenetic and population scales.
    Sites in the alignment containing more than 10\% of gaps across the species were discarded.
    Altogether, our genome-wide dataset contains $14,509$ protein-coding DNA sequences in $87$ placental mammals.

    \subsection{Selection coefficient ($\Sphy$) in a phylogeny-based method}
    \label{subsec:s-phylogeny-method}
    We analyzed the phylogenetic-level data using mutation-selection models.
    These models assume the protein-coding sequences are at mutation-selection balance under a fixed fitness landscape characterized by a fitness vector over the $20$ amino acids at each site~\cite{yang_mutationselection_2008, halpern_evolutionary_1998, rodrigue_mechanistic_2010}.
    Mathematically, the rate of non-synonymous substitution from codon $a$ to codon $b$ ($q_{a \mapsto b}^{(i)}$) at site $i$ of the sequence is equal to the rate of mutation of the underlying nucleotide change ($\mu_{a \mapsto b}$) multiplied by the scaled probability of mutation fixation ($\proba_{a \mapsto b}^{(i)}$).
    The probability of fixation depends on the difference between the scaled fitness of the amino acid encoded by the mutated codon ($F_b^{(i)}$) and the amino acid encoded by the original codon ($F_a^{(i)}$) at site $i$~\cite{wright_evolution_1931, fisher_genetical_1930}.
    The rate of substitution from codon $a$ to $b$ at a site $i$ is thus:
    \begin{equation}
        \begin{dcases}
            q_{a \mapsto b}^{(i)} & = 0 \text{ if codons $a$ and $b$ are more than one mutation away,} \\
            q_{a \mapsto b}^{(i)} & = \mu_{a \mapsto b} \text{ if codons $a$ and $b$ are synonymous, and}\\
            q_{a \mapsto b}^{(i)} & = \mu_{a \mapsto b} \dfrac{F_b^{(i)} - F_a^{(i)}}{1 - \e^{F_a^{(i)} - F_b^{(i)}}} \text{ if codons $a$ and $b$ are non-synonymous}.
        \end{dcases}
    \end{equation}
    Fitting the mutation-selection model on a multi-species sequence alignment leads to an estimation of the $4 \times 4$ nucleotide mutation rate matrix ($\UniDimArray{\mu}$) as well as the $20$ amino-acid fitness landscape ($\UniDimArray{F^{(i)}}$) at each site $i$.
    The selection coefficient for a mutation from codon $a$ to codon $b$ at site $i$ is defined as:
    \begin{equation}
        \Sphy^{(i)} (a \mapsto b) = \Delta F^{(i)} = F^{(i)}_{b} - F^{(i)}_{a}.
    \end{equation}
    In our subsequent derivation the source ($a$) and target ($b$) codons as well as the site ($i$) are implicit and thus never explicitly written.
    We used the Bayesian software \textit{BayesCode} (\url{https://github.com/ThibaultLatrille/bayescode}) to estimate the selection coefficients for each protein-coding gene in the mammalian dataset.
    We ran the Markov chain Monte-Carlo (MCMC) algorithm implemented in BayesCode for $2,000$ generations as described in \textcite{latrille_genes_2023}.
    For each gene, after discarding a burn-in period of $1,000$ generations of MCMC, we obtained posterior mean estimates (over the $1,000$ generations left of MCMC) of the mutation rate matrix ($\UniDimArray{\mu}$) as well as the $20$ amino acid fitness landscape ($\UniDimArray{F^{(i)}}$) at each site $i$.

    \subsection{Polymorphism dataset}
    \label{subsec:polymorphism-dataset}

    The genetic variants representing the population level polymorphisms were obtained from the following species and their available datasets: \textit{Equus caballus} (EquCab2 assembly in the EVA study PRJEB9799~\cite{alabri_whole_2020}), \textit{Bos taurus} (UMD3.1 assembly in the NextGen project: \url{https://projects.ensembl.org/nextgen/}), \textit{Ovis aries} (Oar\_v3.1 assembly in the NextGen project), \textit{Capra hircus} (CHIR1 assembly in the NextGen project), converted to ARS1 assembly with dbSNP identifiers~\cite{sherry_dbsnp_2001}), \textit{Chlorocebus sabaeus} (ChlSab1.1 assembly in the EVA project PRJEB22989~\cite{svardal_ancient_2017}), \textit{Homo sapiens} (GRCh38 assembly in the 1000 Genomes Project~\cite{zheng-bradley_alignment_2017}).
    In total, we analyzed 28 populations across the 6 different species with polymorphism data.
    The data was processed as described in \textcite{latrille_genes_2023}.

    Only bi-allelic single nucleotide polymorphisms (SNPs) found within a gene were in our polymorphism dataset, while nonsense variants and indels were discarded.
    To construct the dataset, we first recovered the location of each SNP (represented by its chromosome, position, and strand) in the focal species and matched it to its corresponding position in the coding sequence (CDS) using gene annotation files (GTF format) downloaded from Ensembl (\url{ensembl.org}).
    We then verified that the SNP downloaded from Ensembl matched the reference in the CDS in FASTA format.
    Next, the position in the CDS was converted to the corresponding position in the multi-species sequence alignment (containing gaps) from the OrthoMaM database (see section~\ref{subsec:s-phylogeny-method}) for the corresponding gene by doing a global pairwise alignment (Biopython function pairwise2).
    This conversion from genomic position to alignment position was only possible when the assembly used for SNP-calling was the same as the one used in the OrthoMaM alignment, the GTF annotations, and the FASTA sequences.
    SNPs were polarized using the three closest outgroups found in the OrthoMaM alignment with est-usfs v2.04~\cite{keightley_inferring_2018}, and SNPs with a derived probability lower than 0.99 were discarded (1\% of false positive).

    \subsection{Mutational opportunities}
    \label{subsec:nunber-of-sites}
    The mutational opportunities of any new mutation refer to its likelihood of falling into a specific category (synonymous, deleterious, nearly-neutral, or beneficial).
    Deriving such opportunities is necessary to estimate the strength of selection exerted at the population scale since different categories might have different mutational opportunities, and thus polymorphism and divergence need to be corrected accordingly (see sections~\ref{subsec:substitution-mapping-in-the-terminal-branch},~\ref{subsec:s-polymorphism-method}, and~\ref{subsec:precisison_recall}).
    To calculate mutational opportunities, we reconstructed the ancestral genome of each of the 28 populations, which differs from the corresponding species' reference genome since it accounts for the variability present in the specific population.
    So, for each population, we used the polarized SNPs to reconstruct the ancestral DNA sequence, containing only substitutions and no segregating polymorphisms.
    If an SNP was still segregating in the population, we used its ancestral version.

    From the reconstructed ancestral genome at the base of the population genealogy, all possible mutations were computed, weighted by the instantaneous rate of change between nucleotides obtained from the mutation rate matrix ($\UniDimArray{\mu}$, see section~\ref{subsec:s-phylogeny-method}), summing to $\mu_{\text{tot}}$ across the whole genome, and to $\mu_{\text{syn}}$ when restricted to synonymous mutations.
    Finally, the mutational opportunities for synonymous mutations were computed as the total number of sites across the genome ($L_{\text{tot}}$) weighted by the proportion of synonymous mutations among all possible mutations as:
    \begin{equation}
        L_{\text{syn}} = L_{\text{tot}} \frac{\mu_{\text{syn}}}{\mu_{\text{tot}}}.
    \end{equation}

>>>>

% Use only LaTeX2e, calling the article.cls class and 12-point type.

\documentclass[12pt]{article}

% Users of the {thebibliography} environment or BibTeX should use the
% scicite.sty package, downloadable from *Science* at
% http://www.sciencemag.org/authors/preparing-manuscripts-using-latex 
% This package should properly format in-text
% reference calls and reference-list numbers.

\usepackage{scicite}

\usepackage{times}

% The preamble here sets up a lot of new/revised commands and
% environments.  It's annoying, but please do *not* try to strip these
% out into a separate .sty file (which could lead to the loss of some
% information when we convert the file to other formats).  Instead, keep
% them in the preamble of your main LaTeX source file.

% The following parameters seem to provide a reasonable page setup.

\topmargin 0.0cm
\oddsidemargin 0.2cm
\textwidth 16cm 
\textheight 21cm
\footskip 1.0cm

\usepackage{xcolor, colortbl}
\definecolor{LIGHTGREY}{gray}{0.9}
\usepackage[colorlinks=false]{hyperref}
\usepackage{amssymb,amsfonts,amsmath,amsthm,mathtools}
\usepackage{bm}
\usepackage[export]{adjustbox}
\usepackage{bbold}
\usepackage{tabu}
\usepackage{hhline}
\usepackage{multicol,multirow,array}
\usepackage{etoolbox}
\usepackage{booktabs}
\usepackage{makecell}
\usepackage{graphicx}
\usepackage{pdfpages}
\pdfinclusioncopyfonts=1

%The next command sets up an environment for the abstract to your paper.
\newcommand{\UniDimArray}[1]{\bm{#1}}
\newcommand{\der}{\text{d}}
\newcommand{\e}{\text{e}}
\newcommand{\Ne}{N_{\text{e}}}
\newcommand{\proba}{\mathbb{P}}
\newcommand{\pfix}{\proba_{\text{fix}}}
\newcommand{\dn}{d_N}
\newcommand{\ds}{d_S}
\newcommand{\dnds}{\dn / \ds}
\newcommand{\Sphy}{S_{0}}
\newcommand{\SphyDel}{\mathcal{D}_0}
\newcommand{\SphyNeu}{\mathcal{N}_0}
\newcommand{\SphyBen}{\mathcal{B}_0}
\newcommand{\Sphyclass}{x}
\newcommand{\SphyclassAlt}{y}
\newcommand{\given}{\mid}
\newcommand{\Spop}{S}
\newcommand{\SpopDel}{\mathcal{D}}
\newcommand{\SpopNeu}{\mathcal{N}}
\newcommand{\SpopBen}{\mathcal{B}}
\newcommand{\ProbaPopDel}{\proba [ \SpopDel]}
\newcommand{\ProbaPopNeu}{\proba [ \SpopNeu ]}
\newcommand{\ProbaPopBen}{\proba [ \SpopBen ]}
\newcommand{\AdvMean}{\beta_b}
\newcommand{\DelMean}{\beta_d}
\newcommand{\thetaSyn}{\theta_{\text{S}}}

\newenvironment{sciabstract}{%
\begin{quote} \bf}
{\end{quote}}

% Include your paper's title here

\title{Mammalian protein-coding genes exhibit widespread beneficial mutations that are not adaptive}

% Place the author information here.  Please hand-code the contact
% information and notecalls; do *not* use \footnote commands.  Let the
% author contact information appear immediately below the author names
% as shown.  We would also prefer that you don't change the type-size
% settings shown here.

\author
{T. {Latrille}$^{1\dag\ast}$, J. {Joseph}$^{2\dag}$, D.~A. {Hartasánchez}$^{1}$, N. {Salamin}$^{1}$\\
\\
\normalsize{$^{1}$Department of Computational Biology, Université de Lausanne, Lausanne, Switzerland}\\
\normalsize{$^{2}$Laboratoire de Biométrie et Biologie Evolutive, UMR5558, Université Lyon 1, Villeurbanne, France}\\
\normalsize{$^{\dag}$These authors contributed equally to this work}\\
\normalsize{$^\ast$To whom correspondence should be addressed; E-mail:  \href{mailto:thibault.latrille@ens-lyon.org}{thibault.latrille@ens-lyon.org}.}
}

% Include the date command, but leave its argument blank.

\date{}

%%%%%%%%%%%%%%%%% END OF PREAMBLE %%%%%%%%%%%%%%%%

\begin{document}

% Double-space the manuscript.

\baselineskip24pt

% Make the title.

\maketitle

% Place your abstract within the special {sciabstract} environment.

\begin{sciabstract}
    Mutations can be beneficial by bringing innovation to their bearer, allowing them to adapt to environmental change.
    However, mutations can also be beneficial because they are reverting previous deleterious changes, simply restoring pre-existing functions.
    By integrating phylogenomic and population data in mammals, we estimated the contribution of beneficial back-mutations to molecular evolution, which had remained widely overlooked so far.
    Applying a mutation-selection model, we estimated amino acid fitness landscapes for mammalian protein-coding genes.
    Assuming these fitness landscapes do not change, any beneficial mutation is necessarily a back-mutation.
    We confirmed that these beneficial back-mutations are positively selected in extant populations and demonstrated that a substantial part of ongoing positive selection is not driven by adaptation to environmental change.
\end{sciabstract}

\section*{Introduction}
Adaptation is one of the main processes shaping the diversity of forms and functions across the tree of life~\cite{darwin_origin_1859}.
Evolutionary adaptation is tightly linked to environmental change and species responding to this change~\cite{merrell_adaptive_1994, gavrilets_adaptive_2009}.
For adaptation to occur, there must be variation within populations, which mostly appears via mutations in the DNA sequence.
While neutral mutations will not impact an individual's fitness, deleterious mutations have a negative effect, and beneficial mutations improve their bearer's fitness.
A beneficial mutation is thus more likely than a neutral mutation to invade the population and reach fixation, resulting in a substitution at the species level.
Upon environmental change, because adaptive beneficial mutations toward new fitness optima are more likely, the number of substitutions also increases (fig.~\ref{fig:fitness-landscape}A).
An increased substitution rate is thus commonly interpreted as a sign of adaptation~\cite{mcdonald_adaptative_1991, smith_adaptive_2002, welch_estimating_2006}.
The availability of large-scale genomic data and the development of theoretical models have enabled the detection and quantification of substitution rate changes across genes and lineages~\cite{yang_statistical_2000, eyre-walker_genomic_2006, moutinho_variation_2019}.
These approaches, now common practice in evolutionary biology, helped better understand the processes underpinning the rates of molecular evolution.
However, a collateral effect has been conflating beneficial mutations with adaptive evolution when adaptive evolution is not the only process that can lead to beneficial mutations~\cite{charlesworth_other_2007, mustonen_fitness_2009}.

@@

\subsection*{Fitness landscape reconstruction}
The mutation-selection framework permits to link the patterns of substitution along a phylogeny with the underlying fitness landscape~\cite{halpern_evolutionary_1998, mccandlish_modeling_2014}.
Such mutation-selection models applied to protein-coding DNA sequences allow us to estimate relative fitnesses for all amino acids for each site of the sequence, explicitly assuming that the underlying fitness landscape is stable along the phylogenetic tree~\cite{rodrigue_mechanistic_2010, tamuri_estimating_2012, rodrigue_detecting_2017}.
Importantly, because mutation-selection models at the phylogenetic scale are based on population-genetics equations, their estimates of selection coefficients are directly interpretable as fitness effects for an individual within a population.

@@

This result further indicated that using $\dnds$ as an estimate of adaptation is biased due to the presence of beneficial back-mutations among the non-synonymous substitutions.
By discarding all beneficial back-mutations we can obtain an estimate of $\dnds$ which is not inflated.
By comparing these two ways of calculating $\dnds$ (see section~1.5 in Materials \& Methods), we calculated that beneficial back-mutations inflate $\dnds$ values by between 9 and 11\% across genera while only representing between 1.2 and 1.4\% of non-synonymous mutations (table~S2-3 and fig.~\ref{fig:homo-afr-results}A for humans).

@@

In addition to clinical prognosis, frequencies at which SNPs are segregating within populations provide information on their selective effects.
For instance, deleterious SNPs usually segregate at lower frequencies because of purifying selection, which tends to remove them from the population (fig.~\ref{fig:homo-afr-results}C for humans).
By gathering information across many SNPs, it is possible to estimate the distribution of fitness effects (DFE) at the population scale, taking synonymous SNPs as a neutral expectation~\cite{eyre-walker_distribution_2006, eyre-walker_estimating_2009, galtier_adaptive_2016, tataru_inference_2017}.
From the estimated DFE, we can derive the proportion of beneficial mutations ($\ProbaPopBen$), nearly-neutral mutations ($\ProbaPopNeu$) and deleterious mutations ($\ProbaPopDel$) at the population scale (see section~1.6 in Materials \& Methods).
These approaches offer a unique opportunity to contrast selection coefficients estimated at the population scale ($\Spop$) and at the phylogenetic scale ($\Sphy$).

Across our selection classes ($\SphyDel$, $\SphyNeu$ and $\SphyBen$), one can ultimately estimate the proportion of correct and incorrect predictions, leading to an estimation of \textit{precision} and \textit{recall} (fig.~\ref{fig:method}F-G and section~1.7 in Materials \& Methods).
Across 28 populations of different mammal species, mutations predicted to be deleterious at the phylogenetic scale ($\ProbaPopDel$) were indeed purged at the population scale, with a \textit{precision} in the range of 89--96\% (table~\ref{table:proba} and fig.~\ref{fig:homo-afr-results}D for humans).
Conversely, a \textit{recall} in the range of 95--100\% implies that mutations found to be deleterious at the population scale were most likely also predicted to be deleterious at the phylogenetic scale (table~\ref{table:proba}).
Altogether, purifying selection is largely predictable and amino acids with negative fitness across mammals have been effectively purged away in each population.

@@

@@

@@

@@

\bibliography{codon_models_bibtex}
\bibliographystyle{Science}

\section*{Acknowledgements}
\label{sec:acknowledgment}
We gratefully acknowledge the help of Mélodie Bastian, Nicolas Lartillot, Carina Farah Mugal, Laurent Duret and Nicolas Gambardella for their advice and reviews concerning this manuscript.
This work was performed using the computing facilities of the CC LBBE/PRABI\@.
This study makes use of data generated by the NextGen Consortium.
\textbf{Funding:}
Université de Lausanne; Agence Nationale de la Recherche, Grant ANR-19-CE12-0019 / HotRec.
The European Union’s Seventh Framework Programme (FP7/2010-2014) provided funding for the project under grant agreement no 244356 - “NextGen”.
\textbf{Author contributions:}
Original idea: T.L.\ and J.J.;
Model conception: T.L., J.J.\ and N.S.;
Code: T.L.;
Data analyses: T.L.\ and J.J.;
Interpretation: T.L., J.J., D.A.H.\ and N.S.;
First draft: T.L.\ and J.J.;
Editing and revisions: T.L., J.J., D.A.H.\ and N.S.
Project management and funding: N.S\@.
\textbf{Competing interests:}
The authors declare no conflicts of interest.
\textbf{Data and materials availability:}
The data underlying this article are available at \url{https://doi.org/10.5281/zenodo.7878954}.
Snakemake pipeline, analysis scripts and documentation are available at \href{https://github.com/ThibaultLatrille/SelCoeff}{github.com/ThibaultLatrille/SelCoeff}.

%Here you should list the contents of your Supplementary Materials -- below is an example. 
%You should include a list of Supplementary figures, Tables, and any references that appear only in the SM. 
%Note that the reference numbering continues from the main text to the SM.
% In the example below, Refs. 4-10 were cited only in the SM.     
\section*{Supplementary materials}
Materials and Methods\\
Supplementary Text\\
Figs. S1 to S16\\
Tables S1 to S3\\
References \textit{(S1-S19)}

\end{document}
